Creating a bushfire dataset \\ An adventure in wrangling spatial data

### 2016-01-27 data_processing.tex
add dataprocessing writeup

% !TeX spellcheck = <none>
\documentclass[11pt,a4paper]{article}
% Change Hons to PhD, MAE, MPhil or MComm as appropraite
% Remove singlespace and singleside for final version.

\usepackage{fullpage,booktabs,amsmath}
\usepackage[section]{placeins}
\usepackage{gensymb,microtype,todonotes}
\usepackage{natbib}
\usepackage{pdfpages}
\usepackage{listings}

%% FLOATS and CAPTIONS
\usepackage{float,caption}
\DeclareCaptionStyle{italic}[justification=centering]
 {labelfont={bf},textfont={it},labelsep=colon}
\captionsetup[figure]{style=italic,format=hang,singlelinecheck=true}
\captionsetup[table]{style=italic,format=hang,singlelinecheck=true}

%% GRAPHICS
\usepackage{graphicx}
\setcounter{topnumber}{2}
\setcounter{bottomnumber}{2}
\setcounter{totalnumber}{4}
\renewcommand{\topfraction}{0.85}
\renewcommand{\bottomfraction}{0.85}
\renewcommand{\textfraction}{0.15}
\renewcommand{\floatpagefraction}{0.7}

\title{Predicting Fire Risk: \\ An Evaluation of the Fire Danger Index}
\author{Elle Saber, Rob J Hyndman \& Di Cook}

@@

\begin{abstract}
  This paper investigates the effectiveness of the Forest Fire Danger Index (FFDI) \citep{mcarthur67} at predicting probability of ignition of bushfire in Victoria using logistic regression, classification trees and random forests. I construct a 10km gridded dataset of Victoria consisting of daily values for FFDI, temperature, wind speed, relative humidity, the drought factor, grassland curing level, vapour pressure at 9am and precipitation between 2000 and 2010. I find that the FFDI alone is a poor predictor of ignition. By adding the input variables to the FFDI, grassland curing, vapour pressure at 9am and precipitation I improve model accuracy and conclude that, of the variables used, grassland curing is the most important in predicting ignition. Combining non-linear regression splines with logistic models produces predictive performance comparable with random forests.
\end{abstract}

\section{Introduction}

In Australia, bushfires are one of the most common natural disasters impacting lives and property. To minimise the damage from fire occurrence, management agencies often allocate resources according to where fires are expected to occur and where fire behaviour could be more severe  \citep{padilla11, wotton05}. In order to plan effectively and manage risks resulting from bushfire, the measures used to predict fire occurrence and severity must be reliable and accurate.

There is limited literature on the effectiveness of the Australian fire danger rating system using statistical techniques, the majority of Australian research uses case studies. Internationally research on fire danger rating systems use standard econometric techniques such as logistic regression or Heckman sample selection models. Machine learning and non-parametric estimation are not used extensively in this area.

This paper uses the Australian Forest Fire Danger index (FFDI) as a baseline model for predicting probability of ignition of bushfire then investigates strategies to improve the accuracy of the predictor. Given the science of combustion and fire behaviour it is highly unlikely that the factors affecting fire ignition impact the probability in a linear manner or without interacting with each other.

I construct a dataset of bushfire ignition and meteorological variables such that predicting fire ignition becomes a binary classification problem. Then I apply both some common machine learning techniques: classification trees and random forests and compare their performance to logistic regression models which use non-parametric smoothing functions. The dataset suffers from severe class imbalance, the number of fires compared to number of non fires is tiny. Accordingly, models are evaluated based on balance accuracy measures and receiver operating characteristic plots.

The greatest improvement in accuracy occurs in models which allow for the input variables to behave non-linearly. By adding spline functions to the logistic model I produce a predictor which is comparable to the random forest in accuracy.

While the FFDI may be a good measure of fire intensity once ignition occurs, it is a poor predictor of fire ignition. Predictor performance is improved with the inclusion of the FFDI input variables, grassland curing, vapour pressure at 9am and precipitation. Specifically of these variables, grassland curing is the most important variable in determining probability of ignition.

\section{Background and Literature Review}

\subsection{Fire Danger Rating Systems and McArthur's Forest Fire Danger Index}
Fire danger rating systems estimate the severity of burning conditions and likelihood of fire over a large area at a particular time \citep{chandler83, andrews03}. The Forest Fire Danger Index (FFDI) \citep{mcarthur67} is the fire danger rating system used in Australia. It integrates weather elements into an index and is used as a key tool in assessing fire danger in Australia. Fire danger rating indices are accepted worldwide for anticipating general fire activity in a given area but less so for evaluating ignition probabilities.

McArthur's FFDI originated as tabulated indices and were subsequently reduced to equations by \citet{noble80} which in turn simplified into the formulation of the FFDI given by
\begin{equation}
  \label{eq FFDI}
  F=2.0 e^{-0.450 + 0.987 \ln(D)-0.0345H+0.0338T+0.0234V},
\end{equation}
where
\begin{itemize}
  \item $D$ is the drought factor index;
  \item $H$ is relative humidity (as a percentage);
  \item $T$ is temperature (degrees Celsius); and
  \item $V$ is average wind velocity (in km/hr).
\end{itemize}

McArthur's FFDI relates weather conditions to expected fire behaviour and rate of spread. It is used in fire fighting operations by weather forecasters and fire agencies to determine fire danger \citep{clarke2013}.

\begin{table}
  \centering
  \begin{tabular}{ll}
    \toprule
    Fire Danger Rating & Forest Fire Danger Index \\
    \midrule
    Low-Moderate       & 0--11 \\
    High               & 12--24 \\
    Very High          & 25--49\\
    Severe             & 50--74\\
    Extreme            & 75--99 \\
    Code Red           & 100+ \\
    \bottomrule
  \end{tabular}
  \caption{The Forest Fire Danger Index and the Fire Danger Rating}
  \label{table:FDR}
\end{table}

While the FFDI has been the most universally accepted fire danger rating system used in Australia for many decades, there are a number of weaknesses which may compromise its ability to accurately predict severe fire.

\begin{enumerate}
  \item Experimental studies conducted during its development largely focused on small scale fires in very specific types of forests and under only moderate weather conditions.
  \item The most severe conditions represented by the FFDI were based on known worst case scenario fires, specifically the 1939 Black Friday fires.
\end{enumerate}

These factors limit the applicability of the FFDI in situations where the weather conditions may be out of these envisioned ranges. This is precisely the situation which occurred on Black Saturday 2009 \citep{harris12}.

There is extensive research on predicting fire behaviour using meteorological and topographical information given ignition occurrence but comparatively little also on  predicting fire occurrence and on evaluating the McArthur index.

\citet{dowdy10} produced an extensive report on Australian fire weather, sensitivity analyses and the relationship between the FFDI and Canadian Fire Weather Index concluding that both indices were most sensitive to temperature and windspeed. Importantly \citet{dowdy10} found in some areas of Australia (Eastern Victoria, Tasmania and south west Western Australia) significant fire activity corresponds with relatively low index values suggesting that the value of the FFDI alone may not be a good predictor of fire activity.

Evaluations of other fire danger indices and weather related indices around the world have used statistical techniques as simple as correlation, and as sophisticated as logistic regression (\cite{andrews03}; \cite{harris14}).

%The only work comparing the FFDI to the FWI is a case study involving a pine plantation fire, the FWI was evaluated in direct comparison to the FFDI  and was found to be the better predictor of extreme fire behaviour \citep{cruz07}. This study will fill this gap in the literature by using statistical analysis to evaluate the performance of the FFDI.

\subsection{Predicting fire activity}

A study linking fire activity to certain climate variables (some of which are inputs to the FFDI, others of which are not) and the El Niño Southern Oscillation (ENSO) index found significant relationships between fire activity and vapour pressure, maximum and minimum temperature and antecedent rainfall \citep{harris14}. Notably the variables which the indices are most sensitive to are different to those found to have the strongest link with fire activity suggesting that the fire danger indices may not contain all the important information for predicting fire activity \citep{harris14}.

Logistic regression is common in studies predicting fire activity. The Bushfire and Natural Hazards Cooperative Research Center used logistic regression over daily historical rasters to predict probability of ignition (Read et al., 2015). Similarly, studies in Europe have used logistic regression with socio-economic factors or geographic features as dependent variables rather than fire danger ratings \citep{del11, zhang13}.

There is some literature from Europe and the U.S. on evaluating Fire Danger Rating systems, specifically the U.S. Fire Danger Rating System and the Canadian Fire Weather Index.  These investigations of fire danger rating systems and fire occurrence have  generally used logistic regression \citep{andrews03, padilla11}.

\section{Data}

The CFA divides the state into different districts for management purposes. I use these districts to capture different terrain types, fuel types and population density in Victoria. Dividing the data into these districts allows us to observe any difference in the behaviour of the variables across different parts of the state. Figure~\ref{fig:dist} shows a map of the districts and their labels.

\begin{figure}
  \centering
  %	\includegraphics[width=1\columnwidth]{figures/Data/cfadis.pdf}
  \caption{Country Fire Authority Districts}
  \label{fig:dist}
\end{figure}

\subsection{Explanatory Variables}

In this thesis, I predict probability of ignition using McArthur's Forest Fire Danger Index (FFDI), the meteorological variables used as inputs for the FFDI, grassland curing, vapour pressure at 9am and rainfall.

\subsubsection{Weather}

The meteorological data set comes from a climate simulation of the Murray Darling Basin using the Weather Research and Forecasting (WFR) model. The WRF Modeling system has been developed by the National Center for Atmospheric Research (NCAR), the National Oceanic and Atmospheric Administrations Centers for Environmental Prediction (NCEP) and the other weather research institutions in the United States. \citet{evans10} used the Advanced Research WFR version~3 \citep{skamarock08} to simulate weather for southeastern Australia at a 10km horizontal grid over 24 years from 1985.

One weakness of this dataset is the fact it is modeled rather than observed. While actual observations would be ideal, there is no observational data at this resolution for this length of time in Australia. \citet{lucas10} developed a historical fire weather data set for 78 Bureau of Meteorology weather stations located around Australia. Not long after, an assessment of quality of this data set by \citet{clarke2013} revealed that only 38 of Lucas's 78 stations could be considered good quality. 38 geographical points over the whole country is too sparse a dataset to be feasible for this project.

The strengths of the dataset make up for its shortcomings. As discussed, the resolution of the WRF model gives a grid resolution not possible from observations. \citet{evans10} evaluate the model output against gridded precipitation and temperature observations available from the Australian Water Availability Project and it was deemed to have successfully reproduced daily statistics capturing spatial patterns  \citep{evans10, clarkeevans13}. An additional advantage of using modelled data is that there are no missing values or false readings from weather stations to create problems in analysis.

The FFDI data is created with the meteorological variables from the WRF model. As per \citet{clarkeevans13} the spatial distribution of the FFDI has also been evaluated against observations with success  \citep{sanabria13}. Clarke and Evans discuss some heterogeneities with modelled windspeed due to changing technology used to measure this variable mostly evident in rural areas before the 1990s resulting in an underestimation of average FFDI values by approximately 5 percent for this period. The fire occurrence dataset begins in 2000 and as a result these windspeed errors do not impact my analysis.

The WRF model \citep{evans10} is deemed to be an accurate representation of Australian weather on this scale and is a useful tool for modelling fire weather over Australia \citep{clarke2013}. The paper uses the FFDI, maximum temperature (Kelvin), wind speed (m/s), relative humidity (\%) and the drought factor calculated by \citet{clarke2013}  from this model. All other variables used in this paper are matched to the 10km grid points in the meteorological dataset.

%Graphs, need to redo better
\begin{figure}
  \centering
  %	\includegraphics[width=1\columnwidth]{figures/Data/silvan.pdf}
  \caption{Time series of meteorological variables for Silvan (-37.82, 145.42)}
  \label{fig:silv}
\end{figure}

Figure~\ref{fig:silv} shows the time series of the meteorological variables for one point in Victoria. The temperature variable is the daily maximum, values are given in Kelvin (\degree C = K - 273.15), and we clearly observe the seasonal cycles with higher temperatures in the summer months and lower temperatures in the winter months. Maximum daily temperatures across the state range from around 2 \degree C to 46\degree  C.

Relative humidity indicates amount of water vapour present in air as a percentage of the amount needed for saturation at the same temperature. Relative humidity at 3pm exhibits the expected seasonal patterns increasing during the winter months and decreasing during the summer months.

Windspeed does not exhibit the same seasonal patterns as the other variables.

The drought factor ranges between 0 and 10 and measures the dryness of the soil using a combination of precipitation, temperature and evaporation calculations \citep{keetch68} clearly exhibits seasonal pattern.

Combination of these variables as per equation (\ref{eq FFDI}) yields the FFDI which clearly exhibits seasonality. Note the spike in summer 2009, corresponding to the Black Saturday bushfires.

%WS, Temp and FDR contours, see if can change colours to match with the diagram
\begin{figure}
  \centering
  %	\includegraphics[width=1\columnwidth]{figures/Data/tw_fdr.pdf}
  \caption{Wind Speed and temperature coloured by fire danger rating}
  \label{fig:tw_fdr}
\end{figure}

The FFDI is most sensitive to temperature and windspeed (\cite{dowdy10}). Figure~\ref{fig:tw_fdr} shows two-dimensional density plots for windspeed and temperature coloured by the fire danger rating (only observations from October--March included) and divided by CFA district. The relative sizes of each contour show the small proportion of extreme and code red days and the difference in weather conditions observed in different parts of Victoria. For instance the only districts observing code red days are districts 16, 17 and 18.

The plots for districts 15 and 20 are missing due to there not being enough data points in each fire dating rating level to produce the 2D densities.

\subsubsection{Grassland Curing}

The level of dying or senescence of plant material caused by seasonal weather patterns is referred to as grassland curing. Grassland curing ranges between 0 and 100, where 0 means the grass is green with high moisture content and 100 means the grass is at its driest. Curing has a significant impact on fire behaviour in grasslands, particularly the probability of ignition and subsequent rate of spread \citep{cheney08}.

The grassland curing data were obtained through the Bureau of Meteorology TERN server and AusCover (http://www.auscover.org.au). AusCover is the remote sensing data products facility of the Terrestrial Ecosystem Research network (TERN, http://www.tern.org.au). There are several different satellite models available to provide curing values for image pixels across the state of Victoria. I use `MapVictoria' \citep{martin15} as this is the current algorithm used by the CFA and is shown to outperform the other curing models.

The curing values are derived from surface reflectance estimates which in turn are derived from cloud-free satellite observations in an 8-day window on a 500-m grid. The dataset spans from the beginning of 2000 to the end of 2010. Missing values may occur in the event of extended cloud cover during the observation window.

Page 10 shows how the level of grassland curing (maximum, minimum and average) changes over each fire season in our sample separated by CFA district. We can see that the districts 2, 5, 8, 16, 17, 18, 20 \&  22 generally dry out faster and have higher minimum curing values relative to districts 4, 6, 9, 10 \& 13. The evolution of grassland curing across the districts clearly demonstrates differing conditions across different parts of the state.

%\includepdf[pages={1}]{figures/Data/curdist.pdf}

% level plot of GS curing, probably needs resizing
\begin{figure}
  \centering
  %	\includegraphics[width=1\columnwidth]{figures/Data/cur_lp.pdf}
  \caption{Level Plot of Grassland Curing }
  \label{fig:cur_lp}
\end{figure}

Figure~\ref{fig:cur_lp} shows a level plot produced using the rasterVis package \citep{rastervis}. We observe grassland curing values across the state for the week beginning 25th January 2009. The horizontal margin shows the mean curing value for each longitude and the vertical margin shows the mean curing for each latitude. The plot clearly shows the difference in curing values across the state, in particular the difference between the east and west of the state. This is consistent with time series graphs of curing.

\subsubsection{Vapour Pressure}

Vapour pressure, in meteorology, describes the partial pressure of water vapour in the atmosphere, it is directly related to the number of water vapour molecules in the air.  \citet{harris14} find there is a significant relationship between fire activity and vapour pressure among other climate variables. I investigate whether vapour pressure at 9am can improve prediction of fire occurrence.

The vapour pressure dataset was also obtained from the Bureau of Meteorology TERN server and AusCover. The vapour pressure dataset is gridded analyses of daily air water vapour pressure from the Bureau of Meteorology's thermometer network across Australia \citep{jones09}, calculated on a 5km grid.

% level plot of vapour pressure, level plot may be unnesc could just have std plot
\begin{figure}
  \centering
  %	\includegraphics[width=1\columnwidth]{figures/Data/vap_ras.pdf}
  \caption{Raster plot of vapour pressure 9am }
  \label{fig:vap_ras}
\end{figure}

Figure~\ref{fig:vap_ras} shows a level plot of the vapour pressure at 9am \citep{rastervis}. The pixel values of the vapour pressure across the state on the 7th of February 2009. We see very clearly in the north-west of the state vapour pressure at 9am is much lower relative to the south-east.

\subsubsection{Precipitation}

The rainfall data are gridded analyses of daily rainfall measurements from the Bureau of Meteorology's rain gauge network (TERN, AusCover) derived from daily rainfall values from weather stations across Australia.

The daily rainfall data represents the amount of precipitation of any type observed by means of rain gauges measuring millimeters of liquid water depth over a 24-hour period. The 24-hour period runs from local time 9am the day before to 9am the current day. Rainfall values from sites across the country are analysed onto a 0.05° × 0.05° grid.

\begin{table}
  \centering
  \begin{tabular}{lrrrrrrrr}
    \toprule
           & ffdi  & temp  & df    & rh    & ws    & curing & rain  & vap \\
    \midrule
    ffdi   & 1.00  & 0.77  & 0.62  & -0.77 & 0.04  & 0.51   & -0.20 & 0.01 \\
    temp   & 0.77  & 1.00  & 0.43  & -0.70 & -0.18 & 0.46   & -0.10 & 0.39 \\
    df     & 0.62  & 0.43  & 1.00  & -0.58 & -0.09 & 0.47   & -0.37 & -0.08 \\
    rh     & -0.77 & -0.70 & -0.58 & 1.00  & 0.19  & -0.49  & 0.31  & 0.06 \\
    ws     & 0.04  & -0.18 & -0.09 & 0.19  & 1.00  & 0.01   & 0.07  & -0.16 \\
    curing & 0.51  & 0.46  & 0.47  & -0.49 & 0.01  & 1.00   & -0.11 & 0.05 \\
    rain   & -0.20 & -0.10 & -0.37 & 0.31  & 0.07  & -0.11  & 1.00  & 0.17 \\
    vap    & 0.01  & 0.39  & -0.08 & 0.06  & -0.16 & 0.05   & 0.17  & 1.00 \\
    \bottomrule
  \end{tabular}
  \caption{Correlation matrix of variables}
  \label{table:correl}
\end{table}

Table~\ref{table:correl} shows the cross correlation matrix for the variables used to predict probability of ignition. High correlation can pose a problem in regression models and to a lesser extent machine learning methods. Variables of concern are temperature, drought factor, relative humidity and FFDI and  which have relatively strong correlation coefficients.

\subsection{Response Variable: Fire Occurrence}

%Commented out until final printing,
%Map of fire ignitions in Victoria
\begin{figure}
  \centering
  %	\includegraphics[width=1\columnwidth]{figures/data/firemap.pdf}
  \caption{Map of fire ignition 2000 - 2009}
  \label{fig:firemap}
\end{figure}

%timeseries histogram with count of ignitions per fortnight
\begin{figure}
  \centering
  %	\includegraphics[width=1\columnwidth]{figures/data/occurence.pdf}
  \caption{Number of ignitions by fortnight}
  \label{fig:occ}
\end{figure}

The fire occurrence data comes from the Fire and Incident Reporting System (FIRS). When the Country Fire Authority or other agency attends a fire or other incident a report is prepared and submitted to the FIRS database. The FIRS dataset used in this study includes all grass, grass and scrub, forest and bushfire in Victoria from the beginning of the year 2000 to 2014.

Recent work by the Bushfire and Natural Hazards Cooperative Research Centre integrating bushfire and planned burn datasets in Victoria give an indication on the reliability of the data (Kilinc et al. 2015).  There is significant uncertainty of accuracy of the fire size, in particular for smaller scale fires. To limit the impact of this imperfection the data is transformed into a binary response variable where a 1 denotes a fire and a 0 denotes no fire.

The data also contains some approximate coordinates for the point of origin for fires, however as discussed below the data is aggregated to a 10km grid. Consequently small errors in the fire coordinates are not deemed to cause a problem.

Figure~\ref{fig:firemap} shows the locations of the fires in the dataset on the map of Victoria. The most noticeable thing in Figure~\ref{fig:firemap} is the sparsity of fires in the north-east and north west of the state and around Melbourne. This is due to the division of fire fighting organisations in Victoria. The sparse areas correspond to National Parks where Parks Victoria or the Department of Environment, Land, Water and Planning (DELWP) coordinate fire fighting activities, so many of the fires occurring in these areas are not included in this dataset.

\begin{figure}
  \centering
  %	\includegraphics[width=1\columnwidth]{figures/Data/tw_fire.pdf}
  \caption{Wind Speed and temperature coloured by fire/no fire}
  \label{fig:tw_fire}
\end{figure}

Figure~\ref{fig:tw_fire} is constructed in the same manner as Figure~\ref{fig:tw_fdr} but instead of grouping by the fire danger rating, I group by Fire/No Fire. Windspeed and temperature play the largest role in determining the FFDI yet there is no discernible difference between the distributions of the two variables when there is a fire compared to no fires. This is the first indicator that the FFDI alone may not be a good predictor of probability of ignition.

\begin{figure}
  \centering
  %	\includegraphics[width=1\columnwidth]{figures/Data/dh_fire.pdf}
  \caption{Drought factor and relative humidity coloured by fire/no fire}
  \label{fig:dh_fire}
\end{figure}

Plotting the 2D densities of the drought factor and relative humidity in Figure~\ref{fig:dh_fire}, ranges of relative humidity don't appear to vary on fire days compared to not fire days but in districts 4, 5, 6, 7, 8, 9, 10, 11, 13, 14 and 24 there is a clear difference in the densities of fire compared to no fire. Small drought factors are mostly associated with non-fire days. In districts 2, 12, 15, 16, 17, 18, 20 and 22 do not have large differences in the two densities, these districts generally have high drought factor and low relative humidity.

\begin{figure}
  \centering
  %	\includegraphics[width=1\columnwidth]{figures/Data/gv_fire.pdf}
  \caption{Grassland curing and vapour pressure coloured by fire/no fire}
  \label{fig:gv_fire}
\end{figure}

In Figure~\ref{fig:gv_fire} shows the 2D densities for grassland curing and vapour pressure for the fire and non-fire days. Once again we see a clear difference in behaviour across districts.

\subsection{Assembling datasets}

@@

The first step was to reduce the dataset to a more manageable size. I converted the NCDF file to a data table using the raster and data.table packages in R \citep{raster, datatable}. I used ESRI shapefiles of Australian Geographical Classification boundaries \citep{ABS1259} to sort the grid points into states and selected only grid points in Victoria. The datasets intersect between 2000 and 2009, subsetting the WFR data reduces the size of the dataset to just over 8 million data points.

From Figure~\ref{fig:occ} it is clear the majority of bushfires occur during the fire season. I reduce the size of the dataset, taking only observations in Victorian fire season (October to March) for a number of reasons. First, from a policy perspective the FFDI, fire danger rating systems, grassland curing are only calculated during the fire season so using these variables outside the time periods they are available does not make intuitive sense. From a practical perspective, this paper is looking at predicting the types bushfires that threaten lives and property which we know to occur in the summer months.

@@

This final dataset is divided into a testing (30\%) and training set (70\%) randomly sampling across `Fire' and `No Fire' such that both datasets have the same proportion of `Fire' (approximately 0.05).

\section{Method}

\subsection{Logistic regression}

Use of logistic models is the usual method for dealing with a binary response problem and it is common in studies on fire occurrence or ignition.  In this project I use logistic regression models and in addition to the linear constructions used in \citet{andrews03} and \citet{del11} I take into account the non-linear relationship believed to exist between the explanatory variables and dependent variable through the inclusion of regression splines.

Regression splines are a method for modelling nonlinear data as a number of piecewise  polynomial functions. The points at which the functions join are called knots. For example when the functions join at $x=k$, $k$ is known as a knot \citep{ruppert03}.

As a result the logistic model becomes
\begin{equation}
  \label{eq logits}
  \Pr(Y=1|X)=\frac{\exp(\sum_{k=1}^K f_k(x_k))}{1+\exp(\sum_{k=1}^K f_k(x_k)}
\end{equation}
where $Y$ denotes fire occurrence, $x_k$ denotes the $k$th predictor and each $f_k$ represents a penalized regression spline.

Regression splines can be subject to over fitting: where the fitted model adheres to the data too strongly and doesn't smooth out `noise' enough. This results in great in-sample predictions but poor out-of-sample predictions. To prevent over fitting occurring we use an automated method that optimizes the number of knots by penalizing for non-smoothness; an overview of using penalized regression splines in this context is given by \citet{wood02}.

\subsection{Classification trees}

Classification tree analysis involves segmenting the predictor space into different regions in order to find the best predictor for the response variable.

The task of finding the single best predictor is achieved using recursive partitioning of the data and can be summarized as two major steps:
\begin{enumerate}
  \item For each predictor multiple cut points are evaluated and the quality of the split is based on the Gini index. Equation (\ref{eq gini}) gives the Gini index for a binary variable:
  \begin{equation}
          \label{eq gini}
          G=\sum_{k=1}^{K} \hat{p}_{mk} (1-\hat{p}_{mk})
  \end{equation}
  where $\hat{p}$ is the estimated probability of a fire at that node. It is also possible to use the cross entropy \citep{james13} which like the Gini index takes on a small value if the node is pure. For each predictor the best cut point is chosen.
  \item The best split overall is chosen:  For each predictor and its corresponding split, the overall best split is chosen again based on Gini index.
\end{enumerate}

The best split for the best predictor defines the first two subsets then same two step procedure is applied to each partition separately and repeated with all subsequent partitions until there is no meaningful reduction in the Gini index \citep{berk08,james13}.

Decision tree analysis results in a model which is simple to explain, interpret and represent graphically \citep{james13}. This is particularly advantageous in the context of fire danger rating systems.
This decision tree is expected to perform well relative to the logistic model as in this situation we expect a simple formula like equation (\ref{eq logits}) to be inadequate in representing the relationship between the predictors and the response variable.

Two common problems associated with decision trees are the problem of overly complex trees (or over fitting) and secondly that in the quest to reduce the Gini index it is possible to subdivide the data so far that there are very few observations at each node.

To approach over fitting, one of two different approaches can be taken: either a rule can be applied that prevents the tree growing too large (for example by setting a threshold on the minimum amount of information gain required to produce a split), or the tree can be allowed to grow large and pruning can be applied afterwards to produce a subtree. In this paper I use 10-fold cross validation on the training data to choose the precision parameter for the classification tree.

Decision trees tend to be very sensitive to the sample used to grow them. Taking this into account I build a number of trees using different balanced samples of the data and compare the output. An alternative solution is random forests (discussed below) as they have the additional advantage of reducing the problem of high variance which plagues decision trees \citep{james13}.

In interpreting classification trees, proximity to the base of the tree indicates how important the variable is in predicting fire occurrence. Variables at the base of the tree are the most important. A tree which does not have the FFDI close to the base of the tree this will suggest the FFDI is not the most important predictor for fire occurrence. Using those variables close to the base of the tree in a fire danger rating system should improve its predictive ability.

\subsection{Random forests}

A random forest is a technique that uses classification trees as a key building block and helps address the limitations of classification trees and improve predictive performance. They involve multiple trees and random samples of our predictors \citep{berk08}.
The general algorithm is as follows for our dataset of N observations and P predictor variables \citep{berk08, varian14}:

\begin{enumerate}
  \item Take a bootstrap sample (random sample of size N with replacement) from the data.
  \item Take a random sample of size $M < P$ (typically $M \approx \sqrt{P}$, \cite{james13}, without replacement) from the predictors.
  \item Construct a classification tree using these samples of data and predictors.
  \item Repeat from step 2 for each subsequent split until the tree is as large as desired. Do not prune.
  \item Drop observations not included in the bootstrap sample from step 1 (known as out-of-bag data) down the tree. The class assigned to each observation is stored along with each observations predictor values.
  \item Repeat steps 1-5 a large number of times to grow a forest of trees.
  \item In order to determine the classification of a new observation. Have each tree make a classification and use a majority vote for the final prediction.
\end{enumerate}

As discussed above, decision trees suffer from high variance. Taking an average of a set of predictions, as we do in random forests, reduces the variance of the predicted values. The second step of the Random Forests algorithm is an integral part of the reducing variance. By taking a sample of M predictors some of the trees will not have a chance to consider the strongest predictors allowing other predictors to add some value and results in less correlated predictions. Averaging many uncorrelated predictions has a greater reduction in variance than averaging many correlated predictions, thus the average prediction for the random forests will be less variable and more reliable \citep{james13}.

The repeated splitting based on parameters by the random forest allows the forest to detect interaction effects between variables. The definition of interaction used for random forests is that two variables interact if a split on one variable in a tree makes a split on the other variable either systematically less possible or more possible \citep{breiman01}. Detection of interactions is based on the gini values,  details of various methods given by \citet{kelly12}. The ability of the forest to detect interaction automatically is a key advantage of the forest over models such as logistic regression where the researcher must specify any interaction terms.
%Klaus: Random forest: one thing you should add is the interaction detection. The repeated splitting based on parameters help the random forest to detect interaction terms in a way automatically. This is a big advantage over logistic regression where you have to specify themselves. The drawback is though, that it does not tell you what the interaction terms are.
% Brieman: The operating definition of interaction used is that variables m and k interact if a split on one variable, say m, in a tree makes a split on k either systematically less possible or more possible. The implementation used is based on the gini values g(m) for each tree in the forest. These are ranked for each tree and for each two variables, the absolute difference of their ranks are averaged over all trees.

Some of the weaknesses of Random Forests is that they can be a black box?in that it is difficult to understand where the output is coming from and they do not offer simple summaries of the relationship between variables in the way decision trees do \citep{varian14}. The advantages of decreased variance and improved accuracy of predictions which random forests bring come at the cost of interpretability, random forests cannot be represented in a clear diagram and it is no longer immediately obvious which variables are the most influential in the procedure \citep{james13}.

I expect the random forest model to give improved out-of-sample fit particularly with highly non-linear data and despite being difficult to interpret, the random forests can offer insight regarding which predictors contribute the biggest improvements to prediction accuracy \citep{varian14}.

\subsection{Classification with unbalanced classes}

Classical machine learning algorithms assume that the response variable has an equal number of observations within each class. Classes are considered imbalanced if the proportion of observations in one class is not equal to the proportion of observations in the other \citep{japkowicz00}. In practice it is very common to have some degree of imbalance and this does not have much effect on the outcome unless the imbalance reaches some threshold.

In the context of fire occurrence, even during the fire season in Victoria instances of bushfire are not that common. For any given day in the dataset at most 4.5\% of the grids experience fire, some days it is possible to have no fires at all. Intuitively, before looking at the data we can expect the proportion of fire days is much smaller than that of non fire days.

Restricting the dataset to the Victorian fire season (October to March) marginally improves the imbalance problem; nevertheless the proportion of fire days of total days for each spatial point is 0.0053. It is safe to say that this situation is one where the degree of imbalance is problematic as a classifier which always assumes `no fire' has an expected error rate of 0.5\% leaving very little room for a more complicated classifier to improve upon.

There are several ways to address the class imbalance problem in machine learning: over sampling, under-sampling, using different splitting criteria \citep{japkowicz00}.

I address the class imbalance problem in two ways, first in building the model I create balanced samples by under sampling the `no fire' class in a similar manner to \citep{padilla11}.  The second approach is to assess each model's performance in predicting the test data using `balanced accuracy' rather than overall accuracy \citep{mosley13, japkowicz00}.

Under-sampling the no fire class addresses the problem of imbalanced classes but I also use it to reduce the serial and spatial correlation that is present in the dataset by using a `near-maximin' sampling method thereby ensuring all the data points in the sample are separate from each from each other in space and time to reduce the impact of correlation.

For consistency all three of the classification techniques are estimated on the balanced training set and evaluated by  on the test data. I evaluate based on receiver operating characteristic (ROC) plots  and using $2\times2$ confusion matrices.

>>>>

\documentclass[11pt,a4paper]{article}

\usepackage{float,fullpage}
\usepackage{gensymb}
\usepackage{natbib}
\usepackage{pdfpages}
\usepackage{listings}

\title{Creating a bushfire dataset \\ An adventure in wrangling spatial data}
\author{Elle Saber, Rob J Hyndman \& Di Cook}

@@

	\section{The Weather and Research Forecasting Model}

	\begin{table}
		\centering
		\caption{My caption}
		\label{my-label}
		\begin{tabular}{|l|l|l|}
			\hline
			Variables                                                       & Source                                                                         & Datatype \\ \hline
			FFDI, temperature, windspeed, relative humidity, drought factor & Weather \& Research Forecasting (WRF) Model                                    & NetCDF   \\ \hline
			Bushfire occurence                                              & Country Fire Authority (CFA) Fire \& Incident Reporting System (FIRS) Database & CSV      \\ \hline
			Grassland Curing                                                & CFA Grassland Curing Database                                                  & Raster   \\ \hline
			Vapour Pressure (9am) \& Precipitation                          & Bureau of Meteorology                                                          & NetCDF   \\ \hline
		\end{tabular}
	\end{table}

@@

	We begin with the temperature NCDF file (although any of them will do) and extract(?right word?) the spatial coordinates from it. The file contains 155 $\times$ 185  grid points covering most of Eastern Australia (QLD, NSW \& VIC). The spatial points are stored as both north-south/east-west coordinates or as longitude and latitudes. Using the State boundaries as given by the Australia Bureau of Statistics ESRI shape files we find the spatial intersection of the Australian States and the WRF model coordinates.

	Given the computational burden involved with shapefiles we only use it once to create a dataframe containing the WRF model coordinates which fall in Victoria. Then after extracting the variables from the NCDF and converting to long format we use dplyr to filter out only the Victorian coordinates.

	The next step is to select a smaller time period. The fire dataset and WRF model intersect between 2000 and 2009 so we subset the WRF model to this time period, reducing the size of the dataset to just over 8 million data points.

 \sec

@@

	This final dataset is divided into a testing (30\%) and training set (70\%) randomly sampling across 'Fire' and 'No Fire' such that both datasets have the same proportion of 'Fire' (approximately 0.05).

		From Figure \ref{fig:occ} it is clear the majority of bushfires occur during the fire season. I reduce the size of the dataset, taking only observations in Victorian fire season (October to March) for a number of reasons. First, from a policy perspective the FFDI, fire danger rating systems, grassland curing are only calculated during the fire season so using these variables outside the time periods they are available does not make intuitive sense. From a practical perspective, this paper is looking at predicting the types bushfires that threaten lives and property which we know to occur in the summer months.
